---
title: "Autoencoding Random Forests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rfae}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is a tutorial for the entire pipeline of the Autoencoding Random Forests (RFAE) algorithm. This consists of:
1. Training a base Random Forests (RF) model
2. Using the trained RF to encode data into embeddings 
3. Decoding the embeddings back into the input space. 

# Random Forest Training

Rather than being trained end-to-end, the RFAE algorithm is more like a post-processing procedure on a trained Random Forest (RF), so first we need to train a RF.

We start with a dataset of interest, and decide whether we want to train the RF supervised ( the class label is not in the autoencoding process), or unsupervised (all features are passed through the pipeline). While the traditional RF algorithm is exclusively used for supervised learning, there have been many variants (such as Adversarial RFs (ARF), Unsupervised RFs (URF), and more) which allow for the class label to be preserved in the learning process, involving either splitting independently of the target, or training the model to distinguish between synthetic generated data and real data.

RFAE is compatible with `ranger`-like objects, which means it can either use the `ranger` RF (supervised by default), or the `adversarial_rf` (unsupervised) from the `arf` package ([Watson et al., 2023](https://proceedings.mlr.press/v206/watson23a.html)).

```{r rfae}
# Load libraries
library(RFAE)
library(data.table)
library(ggplot2)
library(arf)
library(ranger)
set.seed(42)

# Train-test split
trn <- sample(1:nrow(iris), 100)
tst <- setdiff(1:nrow(iris), trn)

# Train RFs, one supervised and one unsupervised
rf <- ranger(Species ~., data = iris[trn, ], num.trees=50)
arf <- adversarial_rf(iris[trn, ], num_trees = 50)
```

We can speed up computations by registering a parallel backend, which can be used by `ranger` and `arf`, as well as the later functions in the pipeline. The default behavior of all functions is to compute in parallel, if possible, so it is advised to do so. How this is done varies across operating systems.

```{r par, eval=FALSE}
# Register cores - Unix
library(doParallel)
registerDoParallel(cores = 2)
```

Windows requires a different setup.

```{r par2, eval=FALSE}
# Register cores - Windows
library(doParallel)
cl <- makeCluster(2)
registerDoParallel(cl)
```

In either case, we can now execute in parallel. 
```{r rfae3}
# Rerun in parallel
arf <- adversarial_rf(iris[trn, ], num_trees=50)
rf <- ranger(Species ~., iris[trn, ], num.trees=50)
```
The result of either functions is an object of class `ranger`, which we can input into the autoencoding pipeline.

# Encoding Stage

The next step is to train the model to encode the data, using `encode`. This function first computes the adjacency matrix between training samples implied by the `rf`. Samples are considered similar if they are routed into the same leaves often throughout the forest. 

The `encode` function requires a `ranger` object, the training data `x` and a parameter `k` for the dimensionality of the embeddings. You can also set `stepsize`, an integer which controls the number of time steps in the diffusion process, and governs whether you want local or global connectivity captured in your embedding. The default value is 1, which gives the most local connectivity; increasing this allows the information captured by the embeddings to be more global. 

The function outputs a list of length 8, containning `Z`, a `k`-dimensional non-linear embedding of `x` implied by `rf`; `A`, the adjacency matrix; and other information from the encoding process to aid with later steps. 

For large datasets, this function can be very computationally expensive, as it requires at least $\mathcal{O}\big(n^2\big)$ space for the adjacency matrix, where $n$ is the sample size.

```{r encoding}
# One encoding for each type of RF
# We choose k=2 to allow visualisation
emap_arf <- encode(arf, iris[trn, ], k=2)
emap_rf <- encode(rf, iris[trn, ], k=2)

# Print out first five rows of embeddings
# The first five rows of x
iris[trn, ][1:5, ]
# The first five embedded samples for ARF
emap_arf$Z[1:5, ]
# The first five embedded samples for RF
emap_rf$Z[1:5, ]
```
We find a slightly different set of embeddings due to the introduction of the Species column into the training stage of the ARF. 

Moving forward in the vignette, we will only use the `emap_arf` encodings for the pipeline to avoid repeating. In practice, the user should decide whether they want to have supervised embeddings (which may be more useful in feature engineering for classification/regression) or unsupervised (which may be more useful in data analysis).

```{r encoding2, fig.height=5, fig.width=8}
# Plot the embedded training data
tmp <- data.frame(
  dim1 = emap_arf$Z[, 1],
  dim2 = emap_arf$Z[, 2],
  class = iris[trn, ]$Species
)
ggplot(tmp, aes(x = dim1, y = dim2, color = class)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    x = "Diffusion Component 1",
    y = "Diffusion Component 2",
    color = "Species"
  )
```

As we expect, the `iris` data is fairly well separated with two components. These encodings by themselves can be used for feature engineering, compression, or clustering, as well as denoising and batch correction - see [Vu et al., 2025](https://arxiv.org/abs/2505.21441) for more details.

In addition to the embeddings, `encode` generates an adjacency matrix `A`, which captures the similarity between training samples as learned by the RF. This is a doubly-stochastic (rows and columns sum to 1) and positive semi-definite matrix. This matrix describes the Random Forest kernel, and can be considered a data-adaptive kernel which works for mixed-type tabular data, and we can consider it for any kernel method.

```{r encoding4}
A <- emap_arf$A
A[1:5, 1:5]
```

With our learned embeddings, which are contained in the `emap` object, we can also project new data onto the principal components, using `predict`. Internally, the function passes the testing data through the RF, which is passed as an argument, computes the adjacency matrix between training and testing samples, then projects data using the Nyst\"om formula, and returns an embedding matrix of size $n_{test} \times k $.

```{r encoding5, fig.height=5, fig.width=8}
# Project testing data
emb <- predict(emap_arf, arf, iris[tst, ])

# Plot test embeddings
tmp <- data.frame(
  dim1 = emb[, 1],
  dim2 = emb[, 2],
  class = iris[tst, ]$Species
)

ggplot(tmp, aes(x = dim1, y = dim2, color = class)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    x = "Diffusion Component 1",
    y = "Diffusion Component 2",
    color = "Species"
  )
```
# Decoding Stage

After finding the embeddings, we can now use `decode_knn` to decode them back to the original input space. Given a test embedding, function works by first finding the `k`-nearest neighbors (kNN) in the embedding space. Then, these samples
are reconstructed using the leaf bounds of the original RF, following the eForest approach described in [Feng & Zhou, 2018](https://ojs.aaai.org/index.php/AAAI/article/view/11732). Finally, the sample is reconstructed by taking a weighted average of the `k`-nearest training samples. 

The `decode_knn` function requires 3 arguments: the original `rf` object used in training, the `z` matrix of embedded data and the `emap` object outputted via `encode`. Optional arguments include `x_tilde`, which is `NULL` by default to tell the function to use the eForest approach, but the exact training data can otherwise be supplied; the `k` parameter for the kNN approach (default k=5), and a `parallel` flag which is enabled by default. 

The function then returns a list consisting of two data frames: `x_hat`, which is the reconstructed data frame, and `x_tilde`, the training samples rebuilt for the decoding process.
```{r decoding1}
# Decode data
out <- decode_knn(arf, emap_arf, emb)
# Reconstructed testing data
out$x_hat[1:5, ]
# Original testing data
iris[tst, ][1:5, ]
```
The reconstructed data in this case is highly similar to the original data on inspection. `decode_knn`'s post-processing also coerces `x_hat` to match the format of the original (i.e., same number of decimal points and same column ordering).

# Reconstruction Error

Of course, beyond inspection, there should be a principled way in which we can evaluate the quality of reconstructed data. Most existing methods for autoencoding (and by extension evaluating) are built for numerical data, so they do not handle mixed-type tabular data in a sensible way - you can one-hot encode data, but what exactly does that mean in terms of column to column $R^2$? 

We provide this in the form of `reconstruction_error`, which takes first the `Xhat` data frame of reconstructed data, and second the original `X`. Note that while `decode_knn` outputs columns in the correct order, this is not actually necessary for `reconstruction_error`. The function calculates prediction error for categorical features, and 1-$R^2$ for numerical features, both of which are between 0 and 1. 

The function then returns a list of 5 elements, with entries for (1) errors for numerical features; (2) errors for categorical features; (3) average numerical error; (4) average categorical error; and (5) the overall average error. 

```{r}
errors <- reconstruction_error(out$x_hat, iris[tst, ])
# Error in numerical features
errors$num_error
# Error in categorical features
errors$cat_error
# Average numerical error
errors$num_avg
# Average categorical error
errors$cat_avg
# Overall error
errors$ovr_error
```

